Gravitational waves from colliding black holes

### 2017-04-10 Report/Draft/draft.tex
Latest, done for the day

\subsection{Parameter estimation}
A signal is described by a total of 16 parameters\cite{props} - time and phase of coalesence $t_c$ and $\phi_c$, two parameters to describe sky location (right ascension, $\alpha$ and declination $\delta$), luminosity distance $D_L$, inclination angle $\iota$ describing the orientation of the binary's total angular momentum with respect to the line of sight, polarisation angle $\psi$, the masses $m_1$, $m_2$, six spin parameters to totally describe the spins on each of the two black holes $\vec{S}_1$, $\vec{S}_2$, and then two eccentricity parameters. The masses and spins of the component black holes are intrinsic parameters which determine the morphology of the waveform, and the remaining are extrinsic parameters. The maximum spin a black hole can have is $m^2$ in natural units, so the convention is to use a dimensionless spin magnitude $a=|\vec{S}|/m^2 \leq 1$. In the first order, the frequency evolution of the 'chirp' signal is approximated by a combination of the masses known as a the chirp mass
\begin{equation}
\mathcal{M}=\frac{(m_1m_2)^{3/5}}{(m_1+m_2)^{1/5}}\propto \bigg(f^{-11/3}\dot{f}\bigg)^{3/5}
\end{equation}
so the specific morphology of the waveform is in some sense determined by a combination of the total mass and the mass ratio.
Given that so many parameters, many of them extrinsic, describe only two sets of timeseries data (one from each detector), parameter estimation is a challenging task and the parameter space is both enormous and wrought with degenerecies. Two of the more thorougly researched degenerecies are those between total mass, distance and inclination, as all three parameters scale the amplitude of the signal and between mass and spin\cite{spindegen}, which is a degenerecy that arises out of post-Newtonian (PN) theory.

@@

Various methods of waveform generation have been attempted, with varying degrees of computational intensity. In general, Post-Newtonian expansions are used in the inspiral phase where the gravitational field is weak enough for approximations to be sufficient. During the merger and ringdown, full numerical relativity simulations are required.

@@

\subsection{Precession and its astrophysical importance}
Considerable research has been done on studying non-spinning binaries and on non-precessing binaries where the spins are aligned or anti-aligned with the orbital angular momentum $\vec{\hat{L}}$\cite{pe3}, but it is only recently that a more complete study of the parameter space has begun\cite{sloos}\cite{pe_latest}. This is partly down to computational resources, as ignoring spin effects leads to a reduced parameter space and less computationally intensive waveforms. There are a unique set of challenges when considering binaries with non-aligned spin, as due to relativistic effects, these binaries precess around the axis of total angular momentum $\vec{\hat{J}}$, giving a time dependence to the orbital plane of the binary\cite{precess1}\cite{precess2}.
\begin{figure}[h]
	\includegraphics[scale=0.75]{fig6.jpg}
	\centering
	\caption{Illustration of a precessing binary system https://phys.org/news/2015-03-insights-black-hole-collisions.html}
	\centering
\end{figure}
This causes the signal in the detector to have an overall amplitude modulation as a function of time due to the change in orientation of the source, and therefore the change in direction of peak emission of GWs. 
\end{figure}

Binary formation models and how estimating chi p is significant for distinguishing bewteen them
\section{Precessive effects on the waveform}
An example of this is shown in Fig. 7, where from this particular perspective the precessive effect appears largest in the case of the in-plane spin being $0.5$, instead of the maximally precessing system with $|\vec{S}|=0.98$. In this case this is due to the fact that the inclination angle is defined with respect to the total angular momentum, and so when we say we are viewing a binary 'edge-on', i.e. at $\iota=\pi/2$, the actual orbital plane of the system will not necessarily be edge-on at all for precessing systems. As a result, it is not always the case that precession affects are most noticeable in systems with the most precession, and a lot depends on the specific combination of source location and the polarisation and inclination angles for a specific event.
\begin{figure}[h]
	\includegraphics[scale=1]{fig7.png}
	\centering
	\caption{Waveforms for three BBH mergers with $m1=30$, $m2=10$ and with spins only in the $x$ direction on the heavier black hole. The inclination is such that the binary is viewed edge-on.}
	\centering
\section{Matching}
Describe matching, and how matches are used to run targeted software injections.
Then identify parameter combinations where we are particularly sensitive to chi p, and those where we are not
\section{Software injections}
Describe software injections
\subsection{Signal extraction}
Data whitening - basically LOSC stuff, how we go from raw data to a signal - move this to intro?
\subsection{Inference pipeline}
Describe the structure of the inference pipeline
\subsection{Inference runs}
Posteriors and discussion of inference results
\subsection{Impact of Virgo}
A look at how Virgo will influence PE, especially chi p estimation
\section{Conclusions and implications}
Summary of results on PE and estimation of chi p, and a prospect on Virgo's impact.
\bibliography{draftbib.bib}

>>>>

\subsection{Parameter estimation}
A signal is described by a total of 16 parameters\cite{props} - time and phase of coalesence $t_c$ and $\phi_c$, two parameters to describe sky location (right ascension, $\alpha$ and declination $\delta$), luminosity distance $D_L$, inclination angle $\iota$ describing the orientation of the binary's total angular momentum with respect to the line of sight, polarisation angle $\psi$, the masses $m_1$, $m_2$, six spin parameters to totally describe the spins on each of the two black holes $\vec{S}_1$, $\vec{S}_2$, and then two eccentricity parameters. The masses and spins of the component black holes are intrinsic parameters which determine the morphology of the waveform, and the remaining are extrinsic parameters. The maximum spin a black hole can have is $m^2$ in natural units, so the convention is to use a dimensionless spin magnitude $a=|\vec{S}|/m^2 \leq 1$. In the first order, the frequency evolution of the 'chirp' signal is approximated by a combination of the masses known as a the chirp mass
\begin{equation}
\mathcal{M}=\frac{(m_1m_2)^{3/5}}{(m_1+m_2)^{1/5}}\propto \bigg(f^{-11/3}\dot{f}\bigg)^{3/5}
\end{equation}
so the specific morphology of the waveform is in some sense determined by a combination of the total mass and the mass ratio. We also define the total mass $M=m_1+m_2$, and the mass ratio $q=m_2/m_1$ adopting the convention that $m1\geq m2$, so $0<q\leq1$.
Given that so many parameters, many of them extrinsic, describe only two sets of timeseries data (one from each detector), parameter estimation is a challenging task and the parameter space is both enormous and wrought with degenerecies. Two of the more thorougly researched degenerecies are those between total mass, distance and inclination, as all three parameters scale the amplitude of the signal and between mass and spin\cite{spindegen}, which is a degenerecy that arises out of post-Newtonian (PN) theory.

@@

Various methods of waveform generation have been attempted, with varying degrees of computational intensity. In general, Post-Newtonian expansions are used in the inspiral phase where the gravitational field is weak enough for approximations to be sufficient. During the merger and ringdown, full numerical relativity simulations are required\cite{imr}.

@@

\subsection{Precession and its astrophysical importance}
Considerable research has been done on studying non-spinning binaries and on non-precessing binaries where the spins are aligned or anti-aligned with the orbital angular momentum $\vec{\hat{L}}$\cite{pe3}, but it is only recently that a more complete study of the parameter space has begun\cite{sloos}\cite{pe_latest}. This is partly down to computational resources, as ignoring spin effects leads to a reduced parameter space and less computationally intensive waveforms. There are a unique set of challenges when considering binaries with non-aligned spin, as due to relativistic effects, these binaries precess around the axis of total angular momentum $\vec{\hat{J}}$, giving a time dependence to the orbital plane of the binary\cite{precess1}\cite{precess2}.
\begin{figure}[h]
	\includegraphics[scale=0.75]{fig6.jpg}
	\centering
	\caption{Illustration of a precessing binary system https://phys.org/news/2015-03-insights-black-hole-collisions.html}
	\centering
\end{figure}
This causes the signal in the detector to have an overall amplitude modulation as a function of time due to the change in orientation of the source, and therefore the change in direction of peak emission of GWs. This can be seen in Fig. 7, where we compare a very slightly precessing system with a maximally precessing one. Due to the computational intensity of dealing with both the generation of waveforms and inference process involving precessing systems, significant efforts have been made to reduce the size of the parameter space using the degenerecies between specific spin combinations to parametrise the spins of a binary. The most successful of these is the adoption of two spin parameters\cite{imr}\cite{chip} that describe the whole binary system, where we replace the six spin parameters with two:
\begin{equation}
\chi_{\text{eff}}=\bigg(\frac{\mathbf{S}_1}{m_1}+\frac{\mathbf{S}_2}{m_2}\bigg)\cdot\frac{\mathbf{\hat{L}}}{M}
\end{equation}
and
\begin{equation}
\chi_\text{p}=\frac{1}{B_1m^2_1}\text{max}(B_1S_{1\perp},B_2S_{2\perp})
\end{equation}
where $B1=2+3/(2q)$ and $B2=2+(3q)/2$. These parameters effectively quantify the amount of in-plane and out of plane spin of the total binary, removing large degenerate portions of the parameter space. As different spin configuations within these parameters are effectively degenerate, no loss of information occurs in this re-parametrisation.
\begin{figure}[h]
	\includegraphics[scale=1]{fig7.png}
	\centering
	\caption{Waveforms for three BBH mergers with $m1=30$, $m2=10$ and with spins only in the $x$ direction on the heavier black hole. The inclination is such that the binary is viewed edge-on.}
	\centering
\end{figure}
The particular focus of this paper is on the challenges of inferring $\chi_p$ in precessing systems. This parameter is of particular importance due to its astrophysical implications. The formation methods of compact binary systems of stellar mass black holes are currently unkown, and a variety of models have been proposed\cite{modal}. Of particular interest is whether the two black holes formed from a common accreting system, or whether the binary was formed by dynamical capture. The former model would imply that the spins on the black holes would generally tend to be aligned with one another and with the orbital angular momentum, however in the dynamical capture model we expect the spins to be more or less uniformly distributed. Given the large number of expected detections over the coming years, a thorough understanding of the detector's response to precessing waveforms and an accurate estimation of our ability to recover $chi_p$ reliably will be crucial to answering these questions, and maximising the scientific yield from this remarkable technology.
\section{Precessive effects on the waveform}
An example of this is shown in Fig. 7, where from this particular perspective the precessive effect appears largest in the case of the in-plane spin being $0.5$, instead of the maximally precessing system with $|\vec{S}|=0.98$. In this case this is due to the fact that the inclination angle is defined with respect to the total angular momentum, and so when we say we are viewing a binary 'edge-on', i.e. at $\iota=\pi/2$, the actual orbital plane of the system will not necessarily be edge-on at all for precessing systems. As a result, it is not always the case that precession affects are most noticeable in systems with the most precession, and a lot depends on the specific combination of source location and the polarisation and inclination angles for a specific event.
\begin{figure}[h]
	\includegraphics[scale=1]{fig8.png}
	\centering
	\caption{Waveforms for three BBH mergers with $m1=30$, $m2=10$ and with spins only in the $x$ direction on the heavier black hole. The inclination is such that the binary is viewed edge-on.}
	\centering
\end{figure}
\section{Matching}
Describe matching, and how matches are used to run targeted software injections.
Then identify parameter combinations where we are particularly sensitive to chi p, and those where we are not
\section{Software injections}
Describe software injections
\subsection{Signal extraction}
Data whitening - basically LOSC stuff, how we go from raw data to a signal - move this to intro?
\subsection{Inference pipeline}
Describe the structure of the inference pipeline
\subsection{Inference runs}
Posteriors and discussion of inference results
\subsection{Impact of Virgo}
A look at how Virgo will influence PE, especially chi p estimation
\section{Conclusions and implications}
Summary of results on PE and estimation of chi p, and a prospect on Virgo's impact.
\bibliography{draftbib.bib}
